A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950, including the Stensele station and the inland

### 2022-02-07 README.Rmd
add time stamp
updates
updated, BP weq

The file *paper.pdf* is the result of this project.

>>>>

Now Bourgeois-Pichat and the log-cube transform are erased, but the present state of
*paper4.pdf* chaotic. No text is updated, only some text is removed.

The file *paper3.pdf* is the old result of the project. Note the time stamp on the front
page, this report is *not* updated for some time.

### 2022-02-07 getdata2.Rmd
updated, BP weq

##Changed 5 feb 2022:
##stentemp <- tempdat("sten", temp_ivl) # Added 27 Jan 2022
source("R/normal.R")
stentemp <- normal(temp_ivl)
## End change

>>>>

##Changed 5 feb 2022:
##stentemp <- tempdat("sten", temp_ivl) # Added 27 Jan 2022
source("R/normal.R") # normal = Norsjö - Malå!!
stentemp <- normal(temp_ivl)
## End change

### 2022-02-07 paper3.Rmd
updated, BP weq

```{r prepare}
source("R/ltx2.R")
infdat$parity <- cut(infdat$parity, c(-2, 1.5, 4.5, 19), labels = c("1", "2-4", "5+"))
infdat$period <- cut(infdat$birthdate, c(1895, 1914, 1935, 1951), dig.lab = 5)
##levels(infdat$subreg) <- c("ume", "ske")
infdat$season <- cut(infdat$week, c(-1, 9.5, 22.5, 35.5, 48.5, 52.5), labels = c("winter", "spring", "summer", "fall", "winter"))
infdat$season <- as.factor(infdat$season)
infdat$emeantemp <- round(infdat$emeantemp)
levels(infdat$socBranch) <- c("office", "farming", "office", "worker", "worker") # NOTE!
```

@@

### Winter and Spring

This period refers to the months *December to May*. A Cox regression involves as
interesting variables *highTemp*, an indicator of temperature at least four degrees
above the expected for at least two weeks in a row, *emeantemp* the expected temperature
the actual weeek, and *aver* the *excess temperature* the actual week.

```{r neoanalw, results = 'asis'}
##library(survival)
winter <- neo[neo$season %in% c("winter", "spring"), ]
winter$season <- factor(winter$season)
##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
##neo$enter <- g(neo$enter * 365)
fit.cr <- coxreg(Surv(enter, exit, event) ~  highTemp + aver + emeantemp + 
                     sex + season + subreg + illeg + parity + socBranch,
                 data = winter)
dr <- drop1(fit.cr, test = "Chisq")
ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, winter and spring. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
     label = "tab:neoanalw", keep = NULL)
```

### Summer and Fall

This period refers to the months *April to October*. A Cox regression involves as
interesting variables *highTemp*, an indicator of temperature at least four degrees
above the expected for at least two weeks in a row, *emeantemp* the expected temperature
the actual weeek, and *aver* the *excess temperature* the actual week.

```{r neoanals, results = 'asis'}
##library(survival)
summer <- neo[neo$season %in% c("summer", "fall"), ]
summer$season <- factor(summer$season)
##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
##neo$enter <- g(neo$enter * 365)
fit.cr <- coxreg(Surv(enter, exit, event) ~  aver + emeantemp + 
                     sex + season + subreg + illeg + parity + socBranch,
                 data = summer)
dr <- drop1(fit.cr, test = "Chisq")
ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, summer and fall. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
     label = "tab:neoanals", keep = NULL)
```

@@

```{r getpneo}
pneo <- age.window(infdat, c(28 / 365, 1))
gpneo <- pneo
gpneo$enter <- g(gpneo$enter * 365)
gpneo$exit <- g(gpneo$exit * 365)
##
oj <- system.time(gtpneo <- toTpch(Surv(enter, exit, event) ~ period + sex + socBranch + illeg + parity + 
                    lowTemp + highTemp + season + aver
                + emeantemp + subreg, 
                data = gpneo, cuts = g(c(28, 365))))
```

@@

### Winter and Spring

```{r postanalw, results='asis'}
winter <- gtpneo[gtpneo$season %in% c("winter", "spring"), ]
winter$season <- factor(winter$season)

fit.tpch <- tpchreg(oe(event, exposure) ~ lowTemp + highTemp + aver + emeantemp + sex + socBranch  + 
                        illeg + parity + subreg, 
                    data = winter, time = age)
dr <- drop1(fit.tpch, test = "Chisq")
ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring. Adjusted for social status and branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanalw", keep = NULL)
```

@@

### Summer and Fall

```{r postanals, results='asis'}
summer <- gtpneo[gtpneo$season %in% c("summer", "fall"), ]
winter$season <- factor(winter$season)

fit.tpch <- tpchreg(oe(event, exposure) ~ lowTemp + highTemp + aver + emeantemp + sex + socBranch  + 
                        illeg + parity + subreg, 
                    data = summer, time = age)
dr <- drop1(fit.tpch, test = "Chisq")
ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring. Adjusted for social status and branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanals", keep = NULL)
```

@@

```
fit.w <- phreg(Surv(enter - 28/365, exit - 28/365, event) ~ 1, data = pneo)
fit <- coxreg(Surv(enter - 28/365, exit - 28/365, event) ~ 1, data = pneo)
```

@@

```{r bpprim, fig.cap = "Postneonatal survival, Weibull vs. Nelson-Aalen."}
fit.w <- phreg(Surv(enter - 28 / 365, exit - 28 / 365, event) ~ 1, data = pneo)
fit <- coxreg(Surv(enter - 28 / 365, exit - 28 / 365, event) ~ 1, data = pneo)
plot(fit, fn = "cum", col = "blue")
xx <- hazards(fit.w)
lines(xx$x, xx$y, col = "red", lty = 2)
```

@@

```
fit.w <- phreg(Surv(enter, exit, event) ~ 1, data = pneo)
fit <- coxreg(Surv(enter, exit, event) ~ 1, data = pneo)
```
Note the difference: No subtraction of the 28 days from the time variables, which implies that the time origin
is *birth* and *not* the 28th day of life. Thus our data are *left truncated* at 28 days, and the 
Weibull attempt estimates neonatal exogeneous mortality through *extrapolation*. The Cox regression approach
is a non-parametric procedure, and thus not capable of any extrapolated estimation, so neonatal
mortality will simply be taken as zero.

```{r ltruw, fig.cap = "Postneonatal survival, truncated Weibull vs. Nelson-Aalen."}
fit.w <- phreg(Surv(enter, exit, event) ~ 1, data = pneo)
fit <- coxreg(Surv(enter, exit, event) ~ 1, data = pneo)
oldpar <- par(mfrow = c(1, 2))
##check.dist(fit, fit.w)
cc <- hazards(fit)
plot(cc[[1]][, 1], cumsum(cc[[1]][, 2]), type = "l", col = "blue", ylim = c(0, 0.07), axes = FALSE,
     xlab = "Age", ylab = "Cumulative hazards")
axis(1, at = c(28 / 365, 1), labels = c("28/365", "1"))
axis(2, las = 1)
box()
xx <- hazards(fit.w)
lines(xx$x, xx$y, col = "red", lty = 2)
plot(cc[[1]][, 1], cumsum(cc[[1]][, 2]), type = "l", col = "blue", ylim = c(0, 0.07), axes = FALSE,
     xlab = "Age", ylab = "Cumulative hazards")
##plot(fit, fn = "cum", col = "blue", ylim = c(0, 0.07), axes = FALSE)
axis(1, at = c(28 / 365, 1), labels = c("28/365", "1"))
axis(2, las = 1)
box()
lines(xx$x, xx$y - xx$y[1], col = "red", lty = 2)
par(oldpar)
```

@@

## Separate exogeneous and endogeneous Infant mortality

@@

```{r crossings, fig.cap = "Biometric analysis.", fig.height = 4}
fit.c <- coxreg(Surv(enter, exit, event) ~ 1, data = infdat)
fit.w <- phreg(Surv(enter, exit, event) ~ 1, data = pneo)
cc <- hazards(fit.c)
plot(c(0, cc[[1]][, 1]), c(0, cumsum(cc[[1]][, 2])), type = "l", col = "blue", axes = FALSE,
     xlab = "Age (days)", ylab = "Cumulative hazards", ylim = c(0, 0.08))
axis(1, at = c(0, 28 / 365, 1), labels = c(0, 28, 365))
axis(2, las = 1, at = round(c(0, max(cumsum(cc[[1]][, 2])), max(xx$y)), 3))
##abline(v = 1)
box()
xx <- hazards(fit.w)
lines(xx$x, xx$y, col = "red", lty = 2)
abline(h = 0, v = 0)
abline(v = 1, lty = 3)
abline(v = 28 / 365, lty = 3)
abline(h = c(max(cumsum(cc[[1]][, 2])), max(xx$y)), lty = 3)
```

>>>>

The *summer* half-year is the weeks 14--39, about 1 April to 30 September, and the *winter*
half-year is the rest, weeks 1-13 and 40--52, 1 January to 31 March and 1 October to 31 December.
This is the division made in @karlsson, and we keep it for comparability reasons.

```{r prepare}
source("R/ltx2.R")
infdat$parity <- cut(infdat$parity, c(-2, 1.5, 4.5, 19), labels = c("1", "2-4", "5+"))
infdat$period <- cut(infdat$birthdate, c(1895, 1914, 1935, 1951), dig.lab = 5)
##levels(infdat$subreg) <- c("ume", "ske")
##infdat$season <- cut(infdat$week, c(-1, 9.5, 22.5, 35.5, 48.5, 52.5), labels = c("winter", "spring", "summer", "fall", "winter"))
##infdat$season <- as.factor(infdat$season)
## New:
infdat$summer <- infdat[infdat$week %in% 14:39, ]
infdat$emeantemp <- round(infdat$emeantemp)
levels(infdat$socBranch) <- c("office", "farming", "office", "worker", "worker") # NOTE!
```

@@

### Winter

This period refers to the months *October to March*. A Cox regression involves as
interesting variables *highTemp*, an indicator of temperature at least four degrees
above the expected for at least two weeks in a row, *emeantemp* the expected temperature
the actual weeek, and *aver* the *excess temperature* the actual week.

```{r neoanalw, results = 'asis'}
##library(survival)
winter <- neo[!neo$summer, ]
##winter$season <- factor(winter$season)
##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
##neo$enter <- g(neo$enter * 365)
fit.cr <- coxreg(Surv(enter, exit, event) ~  highTemp + aver + emeantemp + 
                     sex + subreg + illeg + parity + socBranch,
                 data = winter)
dr <- drop1(fit.cr, test = "Chisq")
ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, October to March. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
     label = "tab:neoanalw", keep = NULL)
```

### Summer

This period refers to the months *April to September*. A Cox regression involves as
interesting variables *highTemp*, an indicator of temperature at least four degrees
above the expected for at least two weeks in a row, *emeantemp* the expected temperature
the actual weeek, and *aver* the *excess temperature* the actual week.

```{r neoanals, results = 'asis'}
##library(survival)
summer <- neo[neo$summer, ]
##summer$season <- factor(summer$season)
##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
##neo$enter <- g(neo$enter * 365)
fit.cr <- coxreg(Surv(enter, exit, event) ~  aver + emeantemp + 
                     sex + subreg + illeg + parity + socBranch,
                 data = summer)
dr <- drop1(fit.cr, test = "Chisq")
ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, April to September. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
     label = "tab:neoanals", keep = NULL)
```

@@

```{r getpneo}
pneo <- age.window(infdat, c(28 / 365, 1))
gpneo <- pneo
gpneo$enter <- g(gpneo$enter * 365)
gpneo$exit <- g(gpneo$exit * 365)
##
oj <- system.time(gtpneo <- toTpch(Surv(enter, exit, event) ~ period + sex + socBranch + illeg + parity + 
                    lowTemp + highTemp + summer + aver
                + emeantemp + subreg, 
                data = gpneo, cuts = g(c(28, 365))))
```

@@

### Winter

```{r postanalw, results='asis'}
winter <- gtpneo[gtpneo$summer == 0, ]
##winter$season <- factor(winter$season)

fit.tpch <- tpchreg(oe(event, exposure) ~ lowTemp + highTemp + aver + emeantemp + sex + socBranch  + 
                        illeg + parity + subreg, 
                    data = winter, time = age)
dr <- drop1(fit.tpch, test = "Chisq")
ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, October to March. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanalw", keep = NULL)
```

@@

### Summer

```{r postanals, results='asis'}
summer <- gtpneo[gtpneo$summer == 1, ]
##winter$season <- factor(winter$season)

fit.tpch <- tpchreg(oe(event, exposure) ~ lowTemp + highTemp + aver + emeantemp + sex + socBranch  + 
                        illeg + parity + subreg, 
                    data = summer, time = age)
dr <- drop1(fit.tpch, test = "Chisq")
ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, April to September. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanals", keep = NULL)
```

@@

```
births <- readRDS("data/births.rds")
pbirths <- age.window(births, c(28 / 365, 1)
fit.w <- phreg(Surv(enter - 28/365, exit - 28/365, event) ~ 1, data = pbirths)
fit <- coxreg(Surv(enter - 28/365, exit - 28/365, event) ~ 1, data = pbirths)
```

@@

```{r bpprim, fig.cap = "Postneonatal survival, Weibull vs. Nelson-Aalen."}
births <- readRDS("data/births.rds")
pbirths <- age.window(births, c(28 / 365, 1))
fit.w <- phreg(Surv(enter - 28 / 365, exit - 28 / 365, event) ~ 1, data = pbirths)
fit <- coxreg(Surv(enter - 28 / 365, exit - 28 / 365, event) ~ 1, data = pbirths)
plot(fit, fn = "cum", col = "blue")
xx <- hazards(fit.w)
lines(xx$x, xx$y, col = "red", lty = 2)
```

@@

```
fit.w <- phreg(Surv(enter, exit, event) ~ 1, data = pbirths)
fit <- coxreg(Surv(enter, exit, event) ~ 1, data = pbirths)
```
Note the difference: No subtraction of the 28 days from the time variables, which implies that the time origin
is *birth* and *not* the 28th day of life. Thus our data are *left truncated* at 28 days, and the 
Weibull attempt estimates neonatal exogeneous mortality through *extrapolation*. The Cox regression approach
is a non-parametric procedure, and thus not capable of any extrapolated estimation, so neonatal
mortality will simply be taken as zero.

```{r ltruw, fig.cap = "Postneonatal survival, truncated Weibull vs. Nelson-Aalen."}
fit.w <- phreg(Surv(enter, exit, event) ~ 1, data = pbirths)
fit <- coxreg(Surv(enter, exit, event) ~ 1, data = pbirths)
oldpar <- par(mfrow = c(1, 2))
##check.dist(fit, fit.w)
cc <- hazards(fit)
plot(cc[[1]][, 1], cumsum(cc[[1]][, 2]), type = "l", col = "blue", ylim = c(0, 0.07), axes = FALSE,
     xlab = "Age", ylab = "Cumulative hazards")
axis(1, at = c(28 / 365, 1), labels = c("28/365", "1"))
axis(2, las = 1)
box()
xx <- hazards(fit.w)
lines(xx$x, xx$y, col = "red", lty = 2)
plot(cc[[1]][, 1], cumsum(cc[[1]][, 2]), type = "l", col = "blue", ylim = c(0, 0.07), axes = FALSE,
     xlab = "Age", ylab = "Cumulative hazards")
##plot(fit, fn = "cum", col = "blue", ylim = c(0, 0.07), axes = FALSE)
axis(1, at = c(28 / 365, 1), labels = c("28/365", "1"))
axis(2, las = 1)
box()
lines(xx$x, xx$y - xx$y[1], col = "red", lty = 2)
par(oldpar)
```

@@

## The biometric analysis of  exogeneous and endogeneous infant mortality

@@

```{r crossings, fig.cap = "Biometric analysis.", fig.height = 4}
fit.c <- coxreg(Surv(enter, exit, event) ~ 1, data = births)
fit.w <- phreg(Surv(enter, exit, event) ~ 1, data = pbirths)
cc <- hazards(fit.c)
plot(c(0, cc[[1]][, 1]), c(0, cumsum(cc[[1]][, 2])), type = "l", col = "blue", axes = FALSE,
     xlab = "Age (days)", ylab = "Cumulative hazards", ylim = c(0, 0.08))
axis(1, at = c(0, 28 / 365, 1), labels = c(0, 28, 365))
axis(2, las = 1, at = round(c(0, max(cumsum(cc[[1]][, 2])), max(xx$y)), 3))
##abline(v = 1)
box()
xx <- hazards(fit.w)
lines(xx$x, xx$y, col = "red", lty = 2)
abline(h = 0, v = 0)
abline(v = 1, lty = 3)
abline(v = 28 / 365, lty = 3)
abline(h = c(max(cumsum(cc[[1]][, 2])), max(xx$y)), lty = 3)
```
